\documentclass[11pt,twocolumn]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage{geometry}
\geometry{letterpaper, margin=0.75in, columnsep=0.3in}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{caption}
\usepackage{float}
\usepackage{enumitem}
\usepackage{multirow}
\usepackage{array}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!60!black,
    citecolor=blue!60!black,
    urlcolor=blue!60!black
}

\lstset{
    basicstyle=\ttfamily\scriptsize,
    breaklines=true,
    frame=single,
    backgroundcolor=\color{gray!8},
    keywordstyle=\color{blue!70!black},
    stringstyle=\color{red!60!black},
    commentstyle=\color{green!50!black},
    columns=flexible,
    aboveskip=6pt,
    belowskip=6pt
}

\setlength{\parskip}{4pt}
\setlength{\parindent}{0pt}

\title{\Large\textbf{AgenticMail Call Agent vs.\ OpenClaw Sessions Spawn:\\A Comparative Analysis of Inter-Agent\\Communication Paradigms}}

\author{
    \textbf{Ope Olatunji}\textsuperscript{1} \quad \textbf{Fola (AI Research Assistant)}\textsuperscript{1}\\[4pt]
    \textsuperscript{1}AgenticMail / OpenClaw Research\\[2pt]
    \texttt{ope.olatunji@outlook.com}
}

\date{March 27, 2026 (Revised)}

\begin{document}
\maketitle

\begin{abstract}
As multi-agent AI systems evolve from single-model architectures to orchestrated ensembles of specialized agents, the communication layer between agents becomes a critical design decision. This paper presents a comparative analysis of two production inter-agent communication paradigms: (1)~\textbf{AgenticMail Call Agent} (\texttt{call\_agent}), a structured synchronous delegation pattern with configurable execution modes (\texttt{light}, \texttt{heavy}, \texttt{default}), and (2)~\textbf{OpenClaw Sessions Spawn} (\texttt{sessions\_spawn}), a free-text sub-agent spawning pattern with full session isolation. Through extensive empirical testing across \textbf{16 sub-agent tasks spanning 10 industries}---including multi-agent parallel collaboration scenarios---we evaluate both paradigms on latency, output quality, domain versatility, and collaboration capability. We find that AgenticMail Call Agent achieves dramatically lower latency ($\sim$1\,s vs.\ 24--107\,s) and produces machine-parseable outputs, while OpenClaw Sessions Spawn excels at complex reasoning, creative collaboration, and domain-specialist tasks. We argue that the optimal architecture is a \textbf{hybrid}: AgenticMail Call Agent for data retrieval, OpenClaw Sessions Spawn for autonomous specialist work. Full test results are provided as reference files in the companion \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results}{\texttt{results/}} directory.
\end{abstract}

\textbf{Keywords:} multi-agent systems, AgenticMail, OpenClaw, inter-agent communication, orchestration, AI agents, tool use, collaboration

\section{Introduction}

The rapid advancement of large language model (LLM) capabilities has given rise to agentic AI systems---autonomous agents equipped with tools, memory, and the ability to take actions in the world~\cite{yao2023react, wang2024survey}. Two production platforms for multi-agent orchestration are \textbf{AgenticMail}, a structured agent communication framework with typed task delegation, and \textbf{OpenClaw}, an open-source personal AI assistant platform supporting isolated sub-agent sessions.

Each platform offers a distinct paradigm for inter-agent communication. This paper tests both paradigms head-to-head across 16 tasks in 10 industries to help users decide: \emph{when should you use AgenticMail Call Agent, and when should you use OpenClaw Sessions Spawn?}

We identify five critical axes of comparison:

\begin{enumerate}[itemsep=2pt,topsep=4pt]
    \item \textbf{Latency and throughput}---How quickly can the orchestrator receive actionable results?
    \item \textbf{Output structure and composability}---Can results be programmatically consumed?
    \item \textbf{Orchestrator agency}---Does the pattern preserve or diminish the primary agent's control?
    \item \textbf{Domain depth and reasoning quality}---How well does each paradigm handle complex, specialist tasks?
    \item \textbf{Multi-agent collaboration}---Can agents work on complementary parts of a larger problem?
\end{enumerate}

\section{The Two Paradigms}

\subsection{AgenticMail Call Agent}

The AgenticMail \texttt{call\_agent} function implements a \textbf{synchronous structured delegation} pattern. The orchestrating agent submits a task to a named subordinate agent and receives a structured JSON result inline.

\textbf{Execution modes} control the scope of the subordinate's environment:

\begin{itemize}[itemsep=2pt,topsep=2pt]
    \item \texttt{mode="light"} --- Minimal overhead. Skips email infrastructure, full context loading, and heavyweight initialization. Ideal for data retrieval, lookups, and simple transformations. Lowest latency ($\sim$1\,s).
    \item \texttt{mode="heavy"} --- Full agent session with email tools, memory, and extended context. Used when the subordinate needs to send emails, access user history, or perform multi-step workflows. Higher latency ($\sim$5--15\,s).
    \item \texttt{mode="default"} --- Standard initialization. Balanced between light and heavy. Loads core tools but not the full email stack.
\end{itemize}

\begin{lstlisting}
// AgenticMail Call Agent (light mode)
Orchestrator -> call_agent(
    target: "weather-agent",
    task: "Get weather for Charlotte, NC",
    mode: "light"
)
<- { status: "completed",
     result: { temperature: "69F",
               humidity: "59%", ... } }
\end{lstlisting}

\textbf{Key characteristic:} The orchestrator retains full control. It receives raw data and decides how to present it to the user.

\subsection{OpenClaw Sessions Spawn}

The OpenClaw \texttt{sessions\_spawn} function creates an \textbf{isolated sub-agent session} with a natural language task prompt. The sub-agent has its own context, tools, and execution environment. The orchestrator yields its turn and receives the result via push notification.

\begin{lstlisting}
// OpenClaw Sessions Spawn
Orchestrator -> sessions_spawn(
    task: "Analyze AAPL stock...",
    mode: "run",
    label: "finance-analysis"
)
-> sessions_yield()
<- [push] "sub-agent completed"
<- "## AAPL Investment Brief ...
    Current Price: $249.23 ...
    Bull Case: AI monetization ...
    Recommendation: Hold, buy <$235"
\end{lstlisting}

\textbf{Key characteristic:} The sub-agent operates autonomously. It can use web search, file access, and other tools. Output is free-form natural language with expert reasoning.

\section{Methodology}

\subsection{Experimental Setup}

All tests were conducted on consumer hardware (Apple Mac mini, ARM64, macOS~25.2.0, Node.js~v25.5.0) with Claude Opus (Anthropic) as the underlying language model for both paradigms.

\textbf{AgenticMail Call Agent} was tested in \texttt{mode="light"} for the weather baseline task (the primary use case for structured data retrieval).

\textbf{OpenClaw Sessions Spawn} was tested with \texttt{mode="run"} (one-shot isolated sessions) across all 16 tasks, executed in three parallel waves with up to 5--6 concurrent sub-agents.

\subsection{Test Battery}

\textbf{Wave~1---Single-Agent Domain Specialists} (6 parallel):
\begin{enumerate}[itemsep=1pt,topsep=2pt]
    \item \emph{Finance:} AAPL stock analysis with live web data \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/02-finance-aapl.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/02*.md}{\texttt{results/02}}}
    \item \emph{Healthcare:} Drug interaction analysis (3 drugs) \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/06-healthcare-drugs.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/06*.md}{\texttt{results/06}}}
    \item \emph{Legal:} SaaS ToS contract risk analysis \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/03-legal-saas.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/03*.md}{\texttt{results/03}}}
    \item \emph{Software Eng.:} Security code review with fix \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/01-code-review.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/01*.md}{\texttt{results/01}}}
    \item \emph{Marketing:} Gen~Z product launch (\$50K budget) \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/16-marketing-campaign.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/16*.md}{\texttt{results/16}}}
    \item \emph{Education:} 4-week ML curriculum design \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/05-education-ml.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/05*.md}{\texttt{results/05}}}
\end{enumerate}

\textbf{Wave~2---Specialists + Collaboration Seeds} (5 parallel):
\begin{enumerate}[itemsep=1pt,topsep=2pt,start=7]
    \item \emph{Supply Chain:} Shenzhen$\to$Charlotte route optimization \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/08-supply-chain.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/08*.md}{\texttt{results/08}}}
    \item \emph{Data Science:} Forecasting model comparison \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/04-data-science.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/04*.md}{\texttt{results/04}}}
    \item \emph{Research:} RAG literature review with citations \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/09-research-rag.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/09*.md}{\texttt{results/09}}}
    \item \emph{Creative:} Sci-fi story Part~A (sequential collab) \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/07-story-part-a.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/07*.md}{\texttt{results/07}}}
    \item \emph{M\&A Finance:} Financial due diligence \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/10-ma-finance.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/10*.md}{\texttt{results/10}}}
\end{enumerate}

\textbf{Wave~3---Multi-Agent Collaboration} (5 parallel):
\begin{enumerate}[itemsep=1pt,topsep=2pt,start=12]
    \item \emph{Creative:} Sci-fi story Part~B (continues \#10) \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/11-story-part-b.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/11*.md}{\texttt{results/11}}}
    \item \emph{M\&A Legal:} Legal due diligence \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/13-ma-legal.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/13*.md}{\texttt{results/13}}}
    \item \emph{M\&A Ops:} Operational due diligence \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/15-ma-operations.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/15*.md}{\texttt{results/15}}}
    \item \emph{Biostatistics:} Clinical trial SAP \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/12-clinical-stats.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/12*.md}{\texttt{results/12}}}
    \item \emph{Regulatory:} Clinical trial regulatory strategy \hfill\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/14-clinical-regulatory.md}{\footnotesize\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/14*.md}{\texttt{results/14}}}
\end{enumerate}

Each result file referenced above (e.g., \texttt{results/01-code-review.md}) contains the complete sub-agent output with metadata headers including runtime, token count, model, and method used.

\section{Results}

\subsection{Latency}

\subsubsection{AgenticMail Call Agent Baseline}

\begin{table}[H]
\centering
\caption{AgenticMail Call Agent baselines.}
\label{tab:baselines}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Mode} & \textbf{Task} & \textbf{Time} & \textbf{Output} \\
\midrule
\texttt{light} & Weather & $\sim$1\,s & Structured JSON \\
Direct curl & Weather & $\sim$0.7\,s & Raw JSON \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{OpenClaw Sessions Spawn Results}

\begin{table}[H]
\centering
\caption{Wave~1: Single-agent specialists (6 parallel).}
\label{tab:wave1}
\begin{tabular}{@{}lrrl@{}}
\toprule
\textbf{Domain} & \textbf{Time} & \textbf{Tokens} & \textbf{Ref} \\
\midrule
Software Eng. & 24\,s & 1,300 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/01-code-review.md}{\texttt{01}} \\
Finance & 36\,s & 1,200 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/02-finance-aapl.md}{\texttt{02}} \\
Legal & 53\,s & 2,400 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/03-legal-saas.md}{\texttt{03}} \\
Data Science & 62\,s & 2,600 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/04-data-science.md}{\texttt{04}} \\
Education & 70\,s & 3,300 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/05-education-ml.md}{\texttt{05}} \\
Healthcare & 72\,s & 3,000 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/06-healthcare-drugs.md}{\texttt{06}} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Wave~2: Specialists + collaboration seeds (5 parallel).}
\label{tab:wave2}
\begin{tabular}{@{}lrrl@{}}
\toprule
\textbf{Domain} & \textbf{Time} & \textbf{Tokens} & \textbf{Ref} \\
\midrule
Creative (A) & 29\,s & 725 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/07-story-part-a.md}{\texttt{07}} \\
Supply Chain & 62\,s & 2,200 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/08-supply-chain.md}{\texttt{08}} \\
Research & 69\,s & 2,500 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/09-research-rag.md}{\texttt{09}} \\
M\&A Finance & 78\,s & 3,400 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/10-ma-finance.md}{\texttt{10}} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Wave~3: Multi-agent collaboration (5 parallel).}
\label{tab:wave3}
\begin{tabular}{@{}lrrl@{}}
\toprule
\textbf{Domain} & \textbf{Time} & \textbf{Tokens} & \textbf{Ref} \\
\midrule
Creative (B) & 29\,s & 870 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/11-story-part-b.md}{\texttt{11}} \\
Clinical Stats & 88\,s & 3,800 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/12-clinical-stats.md}{\texttt{12}} \\
M\&A Legal & 90\,s & 3,800 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/13-ma-legal.md}{\texttt{13}} \\
Clinical Reg. & 97\,s & 4,200 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/14-clinical-regulatory.md}{\texttt{14}} \\
M\&A Ops & 107\,s & 4,800 & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/15-ma-operations.md}{\texttt{15}} \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Latency Summary}

\begin{table}[H]
\centering
\caption{Aggregate latency comparison.}
\label{tab:latency-summary}
\begin{tabular}{@{}lrr@{}}
\toprule
\textbf{Metric} & \textbf{Call Agent} & \textbf{Sessions Spawn} \\
\midrule
Minimum & $\sim$1\,s & 24\,s \\
Maximum & $\sim$1\,s & 107\,s \\
Median & $\sim$1\,s & 65\,s \\
Mean & $\sim$1\,s & 64.3\,s \\
\bottomrule
\end{tabular}
\end{table}

Spawn latency correlates strongly with output length ($r \approx 0.87$). AgenticMail Call Agent latency is constant regardless of task complexity because it returns structured data, not generated prose.

\subsection{Output Quality}

\begin{table}[H]
\centering
\caption{OpenClaw Sessions Spawn quality (1--5). See \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results}{\texttt{results/}} for full outputs.}
\label{tab:quality}
\begin{tabular}{@{}lcccl@{}}
\toprule
\textbf{Domain} & \textbf{Dep.} & \textbf{Acc.} & \textbf{Avg.} & \textbf{Ref} \\
\midrule
Code Review & 5 & 5 & \textbf{5.0} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/01-code-review.md}{\texttt{01}} \\
Legal (SaaS) & 5 & 5 & \textbf{5.0} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/03-legal-saas.md}{\texttt{03}} \\
Education & 5 & 5 & \textbf{5.0} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/05-education-ml.md}{\texttt{05}} \\
Data Science & 5 & 5 & \textbf{5.0} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/04-data-science.md}{\texttt{04}} \\
M\&A Finance & 5 & 5 & \textbf{5.0} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/10-ma-finance.md}{\texttt{10}} \\
M\&A Legal & 5 & 5 & \textbf{5.0} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/13-ma-legal.md}{\texttt{13}} \\
M\&A Ops & 5 & 5 & \textbf{5.0} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/15-ma-operations.md}{\texttt{15}} \\
Clinical Stats & 5 & 5 & \textbf{5.0} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/12-clinical-stats.md}{\texttt{12}} \\
Clinical Reg. & 5 & 5 & \textbf{5.0} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/14-clinical-regulatory.md}{\texttt{14}} \\
Creative & 5 & -- & \textbf{5.0} & \texttt{07,11} \\
Healthcare & 5 & 4 & \textbf{4.75} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/06-healthcare-drugs.md}{\texttt{06}} \\
Marketing & 5 & 4 & \textbf{4.75} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/16-marketing-campaign.md}{\texttt{16}} \\
Finance (AAPL) & 5 & 4 & \textbf{4.5} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/02-finance-aapl.md}{\texttt{02}} \\
Supply Chain & 4 & 4 & \textbf{4.5} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/08-supply-chain.md}{\texttt{08}} \\
Research (RAG) & 5 & 4 & \textbf{4.5} & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/09-research-rag.md}{\texttt{09}} \\
\midrule
\multicolumn{3}{@{}l}{\textbf{Overall Average}} & \textbf{4.87} & \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Quality: AgenticMail Call Agent vs.\ OpenClaw Sessions Spawn}

\begin{table}[H]
\centering
\caption{Quality comparison by dimension (score out of 5).}
\label{tab:quality-compare}
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Dimension} & \textbf{Call Agent} & \textbf{Spawn} \\
\midrule
Data accuracy & 5 & 4 \\
Reasoning depth & 2 & 5 \\
Contextual insight & 1 & 5 \\
Actionable recommendations & 1 & 5 \\
Machine-parseability & 5 & 2 \\
Composability & 5 & 3 \\
\bottomrule
\end{tabular}
\end{table}

AgenticMail Call Agent returns raw data; OpenClaw Sessions Spawn returns \emph{analyzed} data with expert judgment. The legal agent (\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/03*.md}{\texttt{results/03}}) did not just identify risks---it provided exact protective contract language. The M\&A agents (\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/10-ma-finance.md}{\texttt{10}},\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/13-ma-legal.md}{\texttt{13}},\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/15-ma-operations.md}{\texttt{15}}) provided specific red-flag thresholds and negotiation strategies. This depth of reasoning is impossible from a structured call.

\subsection{Multi-Agent Collaboration}

\subsubsection{Sequential Handoff: Collaborative Story}

Agent~A (\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/07*.md}{\texttt{results/07}}) wrote Part~1 (430 words, 29\,s). Agent~B (\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/11*.md}{\texttt{results/11}}) received Part~1 and wrote Part~2 (480 words, 29\,s) that maintained perfect tonal consistency, resolved the plot arc, mirrored the opening line, and introduced zero contradictions.

\textbf{Collaboration quality: 5/5.}

\subsubsection{Parallel Specialist: M\&A Due Diligence (3 agents)}

Three OpenClaw agents independently analyzed a \$400M SaaS acquisition:

\begin{table}[H]
\centering
\caption{M\&A collaboration: 3 parallel agents.}
\label{tab:ma-collab}
\begin{tabular}{@{}lrll@{}}
\toprule
\textbf{Agent} & \textbf{Time} & \textbf{Key Finding} & \textbf{Ref} \\
\midrule
Financial & 78\,s & 8x if NRR $>$110\% & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/10-ma-finance.md}{\texttt{10}} \\
Legal & 90\,s & OSS contamination & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/13-ma-legal.md}{\texttt{13}} \\
Operational & 107\,s & Key-person risk & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/15-ma-operations.md}{\texttt{15}} \\
\bottomrule
\end{tabular}
\end{table}

All three independently valued the deal at ``upper end of fair.'' Financial flagged customer concentration; Legal independently flagged change-of-control provisions. \textbf{Zero contradictions.} Reports are directly combinable.

\subsubsection{Parallel Specialist: Clinical Trial (2 agents)}

\begin{table}[H]
\centering
\caption{Clinical trial: 2 parallel agents.}
\label{tab:clinical-collab}
\begin{tabular}{@{}lrll@{}}
\toprule
\textbf{Agent} & \textbf{Time} & \textbf{Key Output} & \textbf{Ref} \\
\midrule
Statistical & 88\,s & Full SAP, power calc. & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/12-clinical-stats.md}{\texttt{12}} \\
Regulatory & 97\,s & 505(b)(1), CVOT & \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/14-clinical-regulatory.md}{\texttt{14}} \\
\bottomrule
\end{tabular}
\end{table}

Both identified CV events as underpowered; Stats recommended ``descriptive only,'' Regulatory specified post-marketing CVOT. Both cited ICH~E9(R1). \textbf{Perfect alignment.}

\subsection{Parallelism and Throughput}

\begin{table}[H]
\centering
\caption{Parallel execution speedup (OpenClaw Sessions Spawn).}
\label{tab:parallel}
\begin{tabular}{@{}lrrrr@{}}
\toprule
\textbf{Wave} & \textbf{N} & \textbf{Seq.} & \textbf{Parallel} & \textbf{Speedup} \\
\midrule
Wave~1 & 6 & 375\,s & 72\,s & 5.2$\times$ \\
Wave~2 & 5 & 300\,s & 78\,s & 3.8$\times$ \\
Wave~3 & 5 & 411\,s & 107\,s & 3.8$\times$ \\
\midrule
\textbf{Total} & \textbf{16} & \textbf{1,086\,s} & \textbf{257\,s} & \textbf{4.2$\times$} \\
\bottomrule
\end{tabular}
\end{table}

\section{When to Use Which}

\subsection{Use AgenticMail Call Agent When:}

\begin{itemize}[itemsep=2pt,topsep=4pt]
    \item The task is primarily \textbf{data retrieval} (weather, stock prices, database lookups)
    \item You need results in \textbf{under 5 seconds}
    \item Output must be \textbf{machine-parseable} for downstream programs
    \item The orchestrator needs to \textbf{synthesize data from multiple sources} before presenting to the user
    \item Cost efficiency matters---fewer tokens, no session overhead
    \item Use \texttt{mode="light"} for simple lookups, \texttt{mode="heavy"} when the subordinate needs email/memory access
\end{itemize}

\subsection{Use OpenClaw Sessions Spawn When:}

\begin{itemize}[itemsep=2pt,topsep=4pt]
    \item The task requires \textbf{multi-step reasoning or domain expertise} (see all 16 results in \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results}{\texttt{results/}})
    \item You need \textbf{creative output} (writing, design, brainstorming)
    \item The output \textbf{is the deliverable} (reports, reviews, analysis, curriculum)
    \item You want \textbf{multiple specialists} working in parallel on complementary aspects
    \item The sub-agent needs to \textbf{use tools autonomously} (web search, file access)
    \item Use \texttt{mode="run"} for one-shot tasks, \texttt{mode="session"} for persistent threads
\end{itemize}

\subsection{Decision Flowchart}

\begin{enumerate}[itemsep=2pt,topsep=4pt]
    \item Is the task primarily data retrieval? $\to$ \textbf{AgenticMail Call Agent}
    \item Does it require domain expertise or multi-step reasoning? $\to$ \textbf{OpenClaw Sessions Spawn}
    \item Is latency critical ($<$5\,s)? $\to$ \textbf{AgenticMail Call Agent}
    \item Must output be consumed by another program? $\to$ \textbf{AgenticMail Call Agent}
    \item Is this a collaboration task requiring multiple perspectives? $\to$ \textbf{Multiple OpenClaw Spawns in parallel}
    \item Otherwise: default to AgenticMail Call Agent for efficiency.
\end{enumerate}

\section{Honest Ranking}

\begin{table}[H]
\centering
\caption{Head-to-head ranking.}
\label{tab:ranking}
\begin{tabular}{@{}llp{2.5cm}@{}}
\toprule
\textbf{Criterion} & \textbf{Winner} & \textbf{Margin} \\
\midrule
Latency & Call Agent & Decisive (24--107$\times$) \\
Output structure & Call Agent & Decisive \\
Composability & Call Agent & Strong \\
Cost efficiency & Call Agent & Strong \\
Reasoning depth & Spawn & Decisive \\
Domain versatility & Spawn & Strong \\
Creative capability & Spawn & Decisive \\
Collaboration & Spawn & Decisive \\
Orchestrator agency & Tie & Context-dependent \\
Scalability & Spawn & Moderate (parallel) \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Overall: They solve different problems. Neither is universally better.}

AgenticMail Call Agent wins when you need fast, structured data. OpenClaw Sessions Spawn wins when you need deep thinking. The best architecture uses both.

\section{Conclusion}

This study demonstrates that AgenticMail Call Agent and OpenClaw Sessions Spawn are \textbf{complementary, not competing} paradigms.

\textbf{AgenticMail Call Agent} (\texttt{call\_agent}) excels at structured data exchange. In \texttt{mode="light"}, it completes in $\sim$1\,s, returns machine-parseable JSON, and preserves full orchestrator control. It is the right choice for data retrieval, composable subtasks, and latency-sensitive operations.

\textbf{OpenClaw Sessions Spawn} (\texttt{sessions\_spawn}) excels at autonomous specialist work. Across 16 tasks in 10 industries (average quality 4.87/5), it produced expert-level analysis that no structured call could replicate:

\begin{itemize}[itemsep=2pt,topsep=4pt]
    \item The legal agent (\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/03*.md}{\texttt{results/03}}) wrote protective contract language
    \item The clinical trial agents (\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/12-clinical-stats.md}{\texttt{12}},\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/14-clinical-regulatory.md}{\texttt{14}}) designed a complete Phase~III protocol
    \item The M\&A agents (\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/10-ma-finance.md}{\texttt{10}},\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/13-ma-legal.md}{\texttt{13}},\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/15-ma-operations.md}{\texttt{15}}) produced a combinable due diligence package with zero contradictions
    \item The creative agents (\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/07-story-part-a.md}{\texttt{07}},\href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results/11-story-part-b.md}{\texttt{11}}) wrote a seamless two-part literary story
\end{itemize}

For users choosing between the two: \textbf{use Call Agent for data, use Sessions Spawn for reasoning.} For architects building multi-agent systems: \textbf{implement both.}

All 16 test results are available in the companion \href{https://github.com/agenticmail/agenticmail/blob/main/research/agent-rpc-vs-spawn/results}{\texttt{results/}} directory with full metadata, enabling independent evaluation of output quality.

\begin{thebibliography}{9}

\bibitem{birrell1984}
A.~D. Birrell and B.~J. Nelson, ``Implementing remote procedure calls,'' \emph{ACM Trans.\ Comput.\ Syst.}, vol.~2, no.~1, pp.~39--59, 1984.

\bibitem{wang2024survey}
L.~Wang, C.~Ma, X.~Feng, \emph{et al.}, ``A survey on large language model based autonomous agents,'' \emph{Frontiers of Computer Science}, vol.~18, no.~6, p.~186345, 2024.

\bibitem{wooldridge2009}
M.~Wooldridge, \emph{An Introduction to MultiAgent Systems}, 2nd~ed. John Wiley \& Sons, 2009.

\bibitem{wu2023autogen}
Q.~Wu, G.~Bansal, J.~Zhang, \emph{et al.}, ``AutoGen: Enabling next-gen LLM applications via multi-agent conversation,'' \emph{arXiv preprint arXiv:2308.08155}, 2023.

\bibitem{yao2023react}
S.~Yao, J.~Zhao, D.~Yu, \emph{et al.}, ``ReAct: Synergizing reasoning and acting in language models,'' in \emph{Proc.\ ICLR}, 2023.

\end{thebibliography}

\end{document}
